Haplotype Inference via Boolean Satisfiability and Integer Linear Programming

### 2020-11-30 paper.tex
add comments, minor mods

% Graphics
%\usepackage{rotating}
\usepackage{graphicx}
\graphicspath{ {./resources/images/} }
\usepackage{multirow}

@@

Practically efficient algorithms for solving the SAT problem have been developed despite the fact that SAT is an NP-Complete problem~\cite{DBLP:conf/stoc/Cook71}, 
and SAT has become an importent problem solving technique with applications in model checking and verification, planning, scheduling, cryptography, computational biology et cetera.
The most efficient current SAT solvers are based on the \emph{conflict-driven clause learning} algorithm (CDCL) and use many optimizations and efficient data structures~\cite{DBLP:conf/iccad/SilvaS96, DBLP:journals/tc/Marques-SilvaS99, DBLP:conf/dac/MoskewiczMZZM01, DBLP:conf/aaai/GomesSK98, DBLP:journals/dam/GoldbergN07, ryan2004efficient, DBLP:conf/aaai/BayardoS97, DBLP:conf/sat/LewisSB05}.

@@

\section{Haplotype inference}
\label{sec:hipp}
The human genetic information is contained in the double-stranded DNA molecules consisting of a linear sequence of 4 bases (\textbf{A}denosine, \textbf{C}ytosine, \textbf{G}uanine, \textbf{T}hymine) in the phosphate-deoxyribose backbone.
The DNA molecules, which are packed into \emph{chromosomes}, consist of protein-coding intervals, which are called genes, and other non-coding parts.
Different variants of a gene are called \emph{alleles}.
Humans are a \emph{diploid} species, which means that we have two copies of each chromosome (one from each parent).
We thus have two alleles of each gene, and we are said to be \emph{homozygous} with regard to a gene if we have two identical alleles and \emph{heterozygous} otherwise.

@@

The input to the HI problem is the genotype data of $n$ individuals with the genotype vectors being all of length $m$.
The objective is to find a pair of haplotype vectors for each of the $n$ genotypes such that the two haplotype vectors yield exactly the genotype vector.
That is, for each genotype vector $g$ find a pair of haplotype vectors $(h, h')$ such that $g$ corresponds to $(h, h')$.
Without additional constraints the problem is easy to solve but yields no guarantee that the chosen haplotypes are likely to be the correct ones.
The Haplotype Inference by Pure Parsimony (HIPP) sets another constraint on this problem, namely that the set of distinct haplotype vectors used should be as small as possible.
This condition has been experimentally validated to yield relatively accurate haplotype data~\cite{DBLP:journals/bioinformatics/WangX03a}.
The HIPP problem was shown by Lancia et al to be APX-hard~\cite{DBLP:journals/informs/LanciaPR04}.

@@

\section{SAT formulation}
\label{sec:hipp-sat}
Denote by $\mathcal{G}=\{g_i\mid 1\leq i \leq n\}$ a set of $n$ genotype vectors of length $m$ and by $g_{i,j}$ the $j$'th entry of genotype vector $g_i$.
We construct a propositional formula $\phi (\mathcal{G}, r)$ which is satisfiable if and only if there exists a set of haplotypes $\mathcal{H}$ of size $r$ such that each genotype $g_i\in \mathcal{G}$ is explained by some pair of haplotypes $(h_j, h_k)$ in $\mathcal{H}$.
To solve the HIPP problem the satisfiability of $\phi (\mathcal{G}, r)$ is checked for increasing values of $r$ until a value is found such that $\phi (\mathcal{G}, r)$ is satisfiable but $\phi (\mathcal{G}, r-1)$ is not.
This value of $r$ is then the minimum number of haplotypes required, and the actual haplotypes can be extracted from the satisfying assignment to $\phi (\mathcal{G}, r)$.

@@

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{i,j}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{k,i}\rightarrow\neg h_{k,j}$ and $s^b_{k,i}\rightarrow\neg h_{k,j}$ for all $1\leq k \leq r$.
Similarly, if $g_{i,j}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{k,i}\rightarrow h_{k,j}$ and $s^b_{k,i}\rightarrow h_{k,j}$ for all $1\leq k \leq r$.
If $g_{i,j}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{i,j}$ and $g^b_{i,j}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{i,j}\vee g^b_{i,j})\wedge(\neg g^a_{i,j}\vee \neg g^b_{i,j})$ enforce $g^a_{i,j}$ and $g^b_{i,j}$ to take on different values.
The constraints $s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j})$ and $s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j})$ then enforce the selected haplotypes to take on values dictated by $g^a_{i,j}$ and $g^b_{i,j}$.
\begin{figure}
\centering
\small
\begin{align}
\text{for all  }& i\in\{1,\dots , n\}:\nonumber\\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^a_{k,i} = 1 \\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^b_{k,i} = 1 \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=0:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow \neg h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow \neg h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=1:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=2:\nonumber\\
& \quad\quad\quad\quad g^a_{i,j}\vee g^b_{i,j} \\
& \quad\quad\quad\quad \neg g^a_{i,j}\vee \neg g^b_{i,j} \\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j}) \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j}) \\
\end{align}
\caption{The plain HIPP-SAT encoding.}
\label{fig:enc-hipp-sat}
\end{figure}
The encoding is presented in Figure~\ref{fig:enc-hipp-sat}.

@@

\begin{figure}
\centering
Minimize $\sum_{1\leq l \leq L} x_l$
\begin{align}
\text{for all  }i\in\{1,\dots , n\}:&\quad \sum_{1\leq j \leq 2^{k_i-1}} y_{i,j} = 1 \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:& \quad y_{i,j} - x_{l(j,a)}  \leq  0 \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:&\quad y_{i,j} - x_{l(j,b)}  \leq  0 \\
\text{for all  }l\in\{1,\dots , L\}:&\quad x_l \in \{0,1\} \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:&\quad y_{i,j} \in \{0,1\} 
\end{align}
\caption{The RTIP encoding.}
\label{fig:enc-rtip}
\end{figure}

@@

\begin{figure}
\centering
Minimize $\sum_{1\leq i \leq 2n} x_i$
\begin{align}
\text{for all  }i\in\{1,\dots , n\},\ j\in\{1,\dots ,m\}:&\quad h_{2i,j} + h_{2i-1,j} = g_{i,j} \\
\text{for all  }i, j\in\{1,\dots , 2n\},\text{ such that } i<j:&\quad d_{i,j} \geq h_{i,k}-h_{j,k} \\
&\quad d_{i,j} \geq h_{j,k}-h_{i,k} \\
\text{for all  }i\in\{1,\dots , 2n\}:&\quad x_{i} \geq 2 - i +  \sum_{j=1}^{i-1} d_{i,j} \\
\text{for all  }i,j\in\{1,\dots , 2n\},\ k\in\{1,\dots ,m\}:&\quad x_i\in\{0,1\} \\
&\quad h_{i,k}\in\{0,1\} \\
&\quad d_{i,j}\in\{0,1\}
\end{align}
\caption{The PolyIP encoding.}
\label{fig:enc-polyip}
\end{figure}

@@

This structural difference is reflected in the sizes of the encodings.
Let $n$ be the number of genotypes and $m$ the number of loci.
The size of HIPP-SAT scales polynomially in $n$, $m$ and $r$, and the size of PolyIP is bounded by a polynomial in $n$ and $m$.
The size of RTIP, however, scales linearly with respect to $n$, but exponentially with respect to the number of heterozygous loci (which is bounded above by $m$).
That is, HIPP-SAT and PolyIP have polynomial size whereas the size of RTIP scales exponentially.

>>>>

% Graphics
%\usepackage{rotating}
\usepackage{graphicx}
\graphicspath{ {./resources/images/} }
\usepackage{multirow}
\usepackage[linewidth=1pt]{mdframed}

@@

Practically efficient algorithms for solving the SAT problem have been developed despite the fact that SAT is an NP-Complete problem~\cite{DBLP:conf/stoc/Cook71}, 
and SAT has become an important problem solving technique with applications in model checking and verification, planning, scheduling, cryptography, computational biology et cetera.
The most efficient current SAT solvers are based on the \emph{conflict-driven clause learning} algorithm (CDCL) and use many optimizations and efficient data structures~\cite{DBLP:conf/iccad/SilvaS96, DBLP:journals/tc/Marques-SilvaS99, DBLP:conf/dac/MoskewiczMZZM01, DBLP:conf/aaai/GomesSK98, DBLP:journals/dam/GoldbergN07, ryan2004efficient, DBLP:conf/aaai/BayardoS97, DBLP:conf/sat/LewisSB05}.

@@

\section{Haplotype inference}
\label{sec:hipp}
\TODO{How relevant is the 2-allele haplotype inference? How to generalize to several alleles?}

The human genetic information is contained in the double-stranded DNA molecules consisting of a linear sequence of 4 bases (\textbf{A}denosine, \textbf{C}ytosine, \textbf{G}uanine, \textbf{T}hymine) in the phosphate-deoxyribose backbone.
The DNA molecules, which are packed into \emph{chromosomes}, consist of protein-coding intervals, which are called genes, and other non-coding parts.
Different variants of a gene are called \emph{alleles}.
Humans are a \emph{diploid} species, which means that we have two copies of each chromosome (one from each parent).
We thus have two alleles of each gene, and we are said to be \emph{homozygous} with regard to a gene if we have two identical alleles and \emph{heterozygous} otherwise.

@@

The input to the HI problem is the genotype data of $n$ individuals with the genotype vectors being all of length $m$.
The objective is to find a pair of haplotype vectors for each of the $n$ genotypes such that the two haplotype vectors yield exactly the genotype vector.
That is, for each genotype vector $g$ find a pair of haplotype vectors $(h, h')$ such that $g$ corresponds to $(h, h')$.
Without additional constraints the problem is easy to solve but yields no guarantee that the chosen haplotypes are likely to be the correct ones.
The Haplotype Inference by Pure Parsimony (HIPP) sets another constraint on this problem, namely that the set of distinct haplotype vectors used should be as small as possible.

\TODO{How does $n$ affect the accuracy?}

\TODO{Elaborate: "relatively accurate haplotype data"}

This condition has been experimentally validated to yield relatively accurate haplotype data~\cite{DBLP:journals/bioinformatics/WangX03a}.
The HIPP problem was shown by Lancia et al to be APX-hard~\cite{DBLP:journals/informs/LanciaPR04}.
\TODO{Need to mention what the complexity class APX is!}

@@

\section{SAT formulation}
\label{sec:hipp-sat}
\TODO{Clarify what each subscripted variable means to make reading easier.}

Denote by $\mathcal{G}=\{g_i\mid 1\leq i \leq n\}$ a set of $n$ genotype vectors of length $m$ and by $g_{i,j}$ the $j$'th entry of genotype vector $g_i$.
We construct a propositional formula $\phi (\mathcal{G}, r)$ which is satisfiable if and only if there exists a set of haplotypes $\mathcal{H}$ of size $r$ such that each genotype $g_i\in \mathcal{G}$ is explained by some pair of haplotypes $(h_j, h_k)$ in $\mathcal{H}$.
To solve the HIPP problem the satisfiability of $\phi (\mathcal{G}, r)$ is checked for increasing values of $r$ until a value is found such that $\phi (\mathcal{G}, r)$ is satisfiable but $\phi (\mathcal{G}, r-1)$ is not.
This value of $r$ is then the minimum number of haplotypes required, and the actual haplotypes can be extracted from the satisfying assignment to $\phi (\mathcal{G}, r)$.

@@

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{i,j}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{k,i}\rightarrow\neg h_{k,j}$ and $s^b_{k,i}\rightarrow\neg h_{k,j}$ for all $1\leq k \leq r$.
Similarly, if $g_{i,j}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{k,i}\rightarrow h_{k,j}$ and $s^b_{k,i}\rightarrow h_{k,j}$ for all $1\leq k \leq r$.
If $g_{i,j}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{i,j}$ and $g^b_{i,j}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{i,j}\vee g^b_{i,j})\wedge(\neg g^a_{i,j}\vee \neg g^b_{i,j})$ enforce $g^a_{i,j}$ and $g^b_{i,j}$ to take on different values.
The constraints $s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j})$ and $s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j})$ then enforce the selected haplotypes to take on values dictated by $g^a_{i,j}$ and $g^b_{i,j}$.
\begin{figure}
\centering
\small
\begin{mdframed}
\begin{align}
\text{for all  }& i\in\{1,\dots , n\}:\nonumber\\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^a_{k,i} = 1 \\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^b_{k,i} = 1 \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=0:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow \neg h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow \neg h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=1:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=2:\nonumber\\
& \quad\quad\quad\quad g^a_{i,j}\vee g^b_{i,j} \\
& \quad\quad\quad\quad \neg g^a_{i,j}\vee \neg g^b_{i,j} \\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j}) \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j}) \\
\end{align}
\end{mdframed}
\caption{The plain HIPP-SAT encoding.}
\label{fig:enc-hipp-sat}
\end{figure}

The encoding is presented in Figure~\ref{fig:enc-hipp-sat}.

@@

\begin{figure}
\begin{mdframed}
\centering
Minimize $\sum_{1\leq l \leq L} x_l$
\begin{align}
\text{for all  }i\in\{1,\dots , n\}:&\quad \sum_{1\leq j \leq 2^{k_i-1}} y_{i,j} = 1 \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:& \quad y_{i,j} - x_{l(j,a)}  \leq  0 \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:&\quad y_{i,j} - x_{l(j,b)}  \leq  0 \\
\text{for all  }l\in\{1,\dots , L\}:&\quad x_l \in \{0,1\} \\
\text{for all  }i\in\{1,\dots , n\}, j\in\{1,\dots , 2^{k_i-1}\}:&\quad y_{i,j} \in \{0,1\} 
\end{align}
\end{mdframed}
\caption{The RTIP encoding.}
\label{fig:enc-rtip}
\end{figure}

@@

\begin{figure}
\begin{mdframed}
\centering
Minimize $\sum_{1\leq i \leq 2n} x_i$
\begin{align}
\text{for all  }i\in\{1,\dots , n\},\ j\in\{1,\dots ,m\}:&\quad h_{2i,j} + h_{2i-1,j} = g_{i,j} \\
\text{for all  }i, j\in\{1,\dots , 2n\},\text{ such that } i<j:&\quad d_{i,j} \geq h_{i,k}-h_{j,k} \\
&\quad d_{i,j} \geq h_{j,k}-h_{i,k} \\
\text{for all  }i\in\{1,\dots , 2n\}:&\quad x_{i} \geq 2 - i +  \sum_{j=1}^{i-1} d_{i,j} \\
\text{for all  }i,j\in\{1,\dots , 2n\},\ k\in\{1,\dots ,m\}:&\quad x_i\in\{0,1\} \\
&\quad h_{i,k}\in\{0,1\} \\
&\quad d_{i,j}\in\{0,1\}
\end{align}
\end{mdframed}
\caption{The PolyIP encoding.}
\label{fig:enc-polyip}
\end{figure}

@@

This structural difference is reflected in the sizes of the encodings.
Let $n$ be the number of genotypes, $m$ the number of loci and $r$ the number of distinct haplotypes used in HIPP-SAT.
The size of HIPP-SAT scales polynomially in $n$, $m$ and $r$, and the size of PolyIP is bounded by a polynomial in $n$ and $m$.
The size of RTIP, however, scales linearly with respect to $n$, but exponentially with respect to the number of heterozygous loci (which is bounded above by $m$).
That is, HIPP-SAT and PolyIP have polynomial size whereas the size of RTIP scales exponentially.
